---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.path = "man/figures/README-",
  fig.align = "center",
  out.width = "85%"
)
```

# kerasgenerator <img src="man/figures/logo.svg" align="right" height=140/>

`kerasgenerator` package is shipped with some data generator functions, which are designed to make the data generation process for [`keras`](https://keras.rstudio.com) models easier.

## Installation

The package is in early development stage and currently only available via GitHub. You can install the development version using [`install_github()`](http://devtools.r-lib.org/reference/install_github):

```{r, eval=FALSE}

# install devtools first if you haven't
if (!require(devtools)) install.package("devtools")

# install the development version from GitHub
devtools::install_github("bagasbgy/kerasgenerator")

```

## Further Development

* Support for generating data from **file(s) in specified path**
* Data generator for **k-shot learning** model fitting
